```{r}
# Install all the required packages
# install.packages("Rtsne")
# install.packages("ggplot2")

# PCA for genotype data: https://bcm-uga.github.io/pcadapt/articles/pcadapt.html
library(pcadapt)
library(stats)
# tSNE
library(Rtsne)

# Visualization
library(dplyr)
library(ggplot2)
```

```{r vae-instal}
install.packages("keras") # install the keras R package
library(keras)
install_keras(version = "release")  # install the core Keras library and TensorFlow
set.seed(123)

```


```{r PCA-demo 1}
# https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/prcomp 

C <- chol(S <- toeplitz(.9 ^ (0:31))) # Cov.matrix and its root
all.equal(S, crossprod(C))
set.seed(17)
X <- matrix(rnorm(32000), 1000, 32)
Z <- X %*% C  ## ==>  cov(Z) ~=  C'C = S
all.equal(cov(Z), S, tol = 0.08)
pZ <- prcomp(Z, tol = 0.1)

summary(pZ) # only ~14 PCs (out of 32)
# ## or choose only 3 PCs more directly:
# pz3 <- prcomp(Z, rank. = 3)
# summary(pz3) # same numbers as the first 3 above
# stopifnot(ncol(pZ$rotation) == 14, ncol(pz3$rotation) == 3,
#           all.equal(pz3$sdev, pZ$sdev, tol = 1e-15)) # exactly equal typically

# }
# NOT RUN {
## signs are random
require(graphics)
## the variances of the variables in the
## USArrests data vary by orders of magnitude, so scaling is appropriate
# prcomp(USArrests)  # inappropriate
# prcomp(USArrests, scale = TRUE)
prcomp(~ Murder + Assault + Rape, data = USArrests, scale = TRUE)
# plot(prcomp(USArrests))
# summary(prcomp(USArrests, scale = TRUE))
biplot(prcomp(USArrests, scale = TRUE))
# }

```


```{r tSNE iris-demo}
data(iris)
remove_iris_dup <- unique(iris)

# Remove Duplicate data present in iris
remove_iris_dup <- unique(iris)
 
# Forming the matrix for the first four columns 
iris_matrix <- as.matrix(remove_iris_dup[,1:4])

tsne_out <- Rtsne(iris_matrix)
tsne_plot <- data.frame(x = tsne_out$Y[,1], 
                        y = tsne_out$Y[,2])
ggplot(tsne_plot,label=Species) + geom_point(aes(x=x,y=y))

```


```{r MRM-packages}

# Function to check if a package is installed, and install it if not
install_if_missing <- function(package) {
  if (!require(package, character.only = TRUE)) {
    install.packages(package, dependencies = TRUE)
    library(package, character.only = TRUE)
  }
}

# List of required packages
packages <- c("stats", "Rtsne", "tensorflow", "keras", "vegan", "ecodist", 
              "betapart", "ape", "mFD")

library(vegan)
library(ape)
library(ecodist)
library(betapart)
library(tensorflow)
library(mFD)

# check loaded packages
(.packages())
```


```{r mrm-workflow}
# generated using gpt3 lol
# Load necessary libraries
library(vegan)  # For betapart functions
library(stats)  # For dist function

# Example data (replace with your actual data)
pca_data <- matrix(rnorm(100), ncol = 2)  # Example PCA data
tsne_data <- matrix(rnorm(100), ncol = 2)  # Example t-SNE data
vae_data <- matrix(rnorm(100), ncol = 2)  # Example VAE data
eDNA_presence_absence <- matrix(round(runif(100, 0, 1)), ncol = 10)  # Example eDNA presence/absence data

# Compute Euclidean distance matrices
dist_pca <- dist(pca_data, method = "euclidean")
dist_tsne <- dist(tsne_data, method = "euclidean")
dist_vae <- dist(vae_data, method = "euclidean")

# Compute Jaccard's β-diversity matrix
beta_diversity <- betapart::beta.pair(eDNA_presence_absence, method = "jaccard")

# Example of using the distance matrices in MRM (ecodist package)
mrm_result_pca <- MRM(dist_pca, beta_diversity)
mrm_result_tsne <- MRM(dist_tsne, beta_diversity)
mrm_result_vae <- MRM(dist_vae, beta_diversity)

# Perform permutation tests to assess significance
p_value_pca <- mrm_result_pca$pvalue
p_value_tsne <- mrm_result_tsne$pvalue
p_value_vae <- mrm_result_vae$pvalue

# Compare effectiveness based on p-values and coefficients

```

Explanation:
Euclidean Distance: The dist function in R computes pairwise distances between rows of a matrix using various methods ("euclidean", "manhattan", etc.). Here, "euclidean" calculates the straight-line distance between points in the reduced-dimensional space (2D for PCA, t-SNE, and VAE).

Jaccard's β-Diversity: The betapart::beta.pair function calculates pairwise dissimilarities (β-diversity) between rows of a matrix representing presence/absence data (eDNA_presence_absence). Jaccard's β-diversity is appropriate for binary data (0/1) indicating the presence or absence of MOTUs across samples.

Conclusion:
By computing these distance matrices (Euclidean for continuous data and Jaccard's β-diversity for presence/absence data), you can quantitatively compare the effectiveness of PCA, t-SNE, and VAE in preserving the underlying structure of your eDNA and oceanographic data. These distance matrices are essential inputs for methods like Multiple Regression on Distance Matrices (MRM) to evaluate how well each dimensionality reduction technique captures the ecological relationships and patterns in your dataset.








